# Individual Conditional Expectation (ICE) {#ice}

{{< include _setup.qmd >}}

Individual Conditional Expectation (ICE) plots display one line per instance that shows how the instance's prediction changes when a feature changes.

An ICE plot [@goldstein2015peeking] visualizes the dependence of the prediction on a feature for *each* instance separately, resulting in one line per instance of a dataset.
The values for a line (and one instance) can be computed by keeping all other features the same, creating variants of this instance by replacing the feature's value with values from a grid and making predictions with the black box model for these newly created instances.
The result is a set of points for an instance with the feature value from the grid and the respective predictions.
In other words, ICE plots are all the [ceteris paribus curves](#ceteris-paribus) for a dataset in one plot.

A more formal definition:
In ICE plots, for each instance in $\{(x_{S}^{(i)},x_{C}^{(i)})\}_{i=1}^N$ the curve $\hat{f}_S^{(i)}$ is plotted against $x^{(i)}_{S}$, while $x^{(i)}_{C}$ remains fixed.

## Examples

@fig-ice-bike shows ICE plots for the [bike rental prediction](#bike-data).
The underlying prediction model is a random forest.
All curves seem to follow the same course, so there are no obvious interactions.
But we can also explore possible interactions by modifying the ICE plot.

```{r}
#| label: fig-ice-bike
#| fig-cap: "ICE plots of predicted bicycle rentals by temperature, humidity, and windspeed."
#| fig-asp: 0.4
pred.bike = Predictor$new(bike_rf, bike_test)

p1 = FeatureEffect$new(pred.bike, "temp", method = "ice")$plot() +
  scale_x_continuous("Temperature") +
  scale_y_continuous("Predicted bicycle rentals")

p2 = FeatureEffect$new(pred.bike, "hum", method = "ice")$plot() +
  scale_x_continuous("Humidity") +
  scale_y_continuous("") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p3 = FeatureEffect$new(pred.bike, "windspeed", method = "ice")$plot() +
  scale_x_continuous("Windspeed") +
  scale_y_continuous("") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

# Combine plots with reduced spacing
(p1 | p2 | p3) + plot_layout(guides = "collect", widths = c(1, 1, 1))
```


::: {.callout-tip}

# Use color

By coloring the lines based on another feature's value, you can study interactions.

:::

@fig-ice-bike-colored shows again the ICE plot for humidity, with the difference that the lines are now colored by the season.
This shows a couple of things:
First -- and that's not surprising -- different seasons have different "intercepts".
Meaning that, for example, winter days have an in general lower prediction and summer the highest ones, and that's independent of the humidity.
But @fig-ice-bike-colored also shows that the effect of the humidity differs for the seasons:
In winter, and increase in humidity only slightly reduced the predicted number of bike rentals.
And the effect is rather linear.
For summer, the predicted bike rentals stay more or less flat between 20% and 60% relative humidity and above that drop by quite a chunk.
Humidity effects for spring and fall seem to be a mix of the "winter flatness" and the "summer jump".
However, as indicated by the boxplots in @fig-ice-bike-colored, we shouldn't over-interpret very low humidity effects for summer and fall.

```{r}
#| label: fig-ice-bike-colored
#| fig-cap: ICE curves for the random forest predicting bike rentals. Lines are colored by the season. Above the ICE plots are boxplots showing the distributions of humidity per season.
dat = FeatureEffect$new(pred.bike, "hum", method = "ice")$results
bike2 = bike_test
bike2$.id = 1:nrow(bike_test)

dat = merge(bike2[c(".id", "season")], dat, by = ".id")
iceplot = ggplot(dat) +
  geom_line(aes(group=.id, x=hum, y=.value, color=season), alpha=0.5) +
  scale_y_continuous("Predicted bicycle rentals", limits=c(0, NA)) +
  scale_color_viridis(discrete=TRUE) +
  scale_x_continuous("Humidity")

boxplots <- ggplot(bike) + 
  geom_boxplot(aes(y = season, x = hum), width = 0.1) +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) 
#  coord_fixed(3.6)

(boxplots / iceplot)  +
  plot_layout(heights = c(1, 2))

```



Let's go back to the [penguin classification task](#penguins) and see how the prediction of each instance is related to the feature `bill_length_mm`.
We'll analyze a random forest that predicts the probability of a penguin being female given body measurements.
@fig-ice-penguins is a rather ugly ICE plot.
But sometimes that's the reality.
The reason is that the model is rather sure for most penguins and quickly changes between 0 and 1.

```{r}
#| label: fig-ice-penguins
#| fig-cap: ICE plot of P(Adelie) by bill length. Each line represents a penguin.
pred.penguins = Predictor$new(pengu_rf, penguins_test, class = "female")
ice = FeatureEffect$new(pred.penguins, "bill_length_mm", method = "ice")$plot() + 
  scale_color_discrete(guide='none') + 
  scale_y_continuous('P(female)') +
  scale_x_continuous("Bill length in mm")
ice
```


::: {.callout-tip}

# Use transparency

If lines overlap heavily in a boxplot you can try to make them slightly transparent.
If that doesn't help, you may be better off with a [partial dependence plot](#pdp).

:::

## Centered ICE Plot

There's a problem with ICE plots:
Sometimes it can be hard to tell whether the ICE curves differ between individuals because they start at different predictions.
A simple solution is to center the curves at a certain point in the feature and  display only the difference in the prediction to this point.
The resulting plot is called centered ICE plot (c-ICE).
Anchoring the curves at the lower end of the feature is a good choice.
The new curves are defined as:

$$\hat{f}_{cent}^{(i)}=\hat{f}^{(i)}-\mathbf{1}\hat{f}(x^{a},x^{(i)}_{C})$$

where $\mathbf{1}$ is a vector of 1's with the appropriate number of dimensions (usually one or two), $\hat{f}$ is the fitted model and x^a^ is the anchor point.

### Example

Let's have a look at a centered ICE plot for temperature for the bicycle rental prediction:

```{r}
#| label: ice-bike-centered
#| fig-cap: "Centered ICE plots of predicted number of bikes by temperature. The lines show the difference in prediction compared to the prediction with the temperature fixed at its observed minimum."
predictor = Predictor$new(bike_rf, data = bike_test)
ytext1 = sprintf("Different to prediction at temp = %.1f", min(bike$temp))
ice1 = FeatureEffect$new(predictor, feature = "temp", center.at = min(bike$temp), method = "ice")$plot() +
  scale_y_continuous(ytext1)
#ytext2 = sprintf("Different to prediction at hum = %.1f", min(bike$hum))
#ice2 = FeatureEffect$new(predictor, feature = "hum", center.at = min(bike$hum), method = "ice")$plot() +
#  scale_y_continuous(ytext2)
#ytext3 = sprintf("Different to prediction at windspeed = %.1f", min(bike$windspeed))
#ice3 = FeatureEffect$new(predictor, feature = "windspeed", center.at = min(bike$windspeed), method = "ice")$plot() +
#  scale_y_continuous(ytext3)
#(ice1 | ice2 | ice3)
ice1
```

The centered ICE plots make it easier to compare the curves of individual instances.
This can be useful if we do not want to see the absolute change of a predicted value, but the difference in the prediction compared to a fixed point of the feature range.


## Derivative ICE Plot

Another way to make it visually easier to spot heterogeneity is to look at the individual derivatives of the prediction function with respect to a feature.
The resulting plot is called the derivative ICE plot (d-ICE).
The derivatives of a function (or curve) tell you whether changes occur and in which direction they occur.
With the derivative ICE plot, it is easy to spot ranges of feature values where the black box predictions change for (at least some) instances.
If there is no interaction between the analyzed feature $x_S$ and the other features $x_C$, then the prediction function can be expressed as:

$$\hat{f}(x)=\hat{f}(x_S,x_C)=g(x_S)+h(x_C),\quad\text{with}\quad\frac{\delta\hat{f}(x)}{\delta{}x_S}=g'(x_S)$$

Without interactions, the individual partial derivatives should be the same for all instances.
If they differ, it is due to interactions and it becomes visible in the d-ICE plot.
In addition to displaying the individual curves for the derivative of the prediction function with respect to the feature in S, showing the standard deviation of the derivative helps to highlight regions in feature in S with heterogeneity in the estimated derivatives.
The derivative ICE plot takes a long time to compute and is rather impractical.


## Strengths

Individual conditional expectation curves are **intuitive to understand**.
One line represents the predictions for one instance if we vary the feature of interest.

ICE curves can **uncover heterogeneous relationships**.

## Limitations

ICE curves **can only display one feature** meaningfully, because two features would require the drawing of several overlaying surfaces and you would not see anything in the plot.

ICE curves suffer from correlation: If the feature of interest is correlated with the other features, then **some points in the lines might be invalid data points** according to the joint feature distribution.

If many ICE curves are drawn, the **plot can become overcrowded** and you will not see anything.
The solution: Either add some transparency to the lines or draw only a sample of the lines.

In ICE plots it might not be easy to **see the average**.
This has a simple solution:
Combine individual conditional expectation curves with the [partial dependence plot](#pdp).

## Software and Alternatives

ICE plots are implemented in the R packages `iml` [@molnar2018iml] (used for these examples), `ICEbox` [@goldstein2017package], and `pdp`.
Another R package that does something very similar to ICE is `condvis`.
In Python, you can use [PiML](https://selfexplainml.github.io/PiML-Toolbox/_build/html/index.html) [@sudjianto2023piml].

